---
title: "Assignment 10: Data Scraping"
author: "Mara Michel"
output: pdf_document
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)
library(lubridate)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2022 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 Set the URL
webpage <-read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2022')

webpage
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
Water_System_Name <-webpage %>% 
  html_nodes('div+ table tr:nth-child(1) td:nth-child(2)')%>% html_text()

PWSID <-webpage %>% 
  html_nodes('td tr:nth-child(1) td:nth-child(5)')%>% html_text()
  
Ownership <-webpage %>% 
  html_nodes('div+ table tr:nth-child(2) td:nth-child(4)')%>% html_text()
  
Max_Day_Use <-webpage %>% 
  html_nodes('th~ td+ td') %>% html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2022

```{r create.a.dataframe.from.scraped.data}
#4 
max_daily_22_df <- data.frame("Water_System_Name" = rep(Water_System_Name,12),
                              "PWSID" = rep(PWSID,12), 
                              "Ownership"= rep(Ownership,12),
                              "Month" = c('Jan','May','Sep','Feb','Jun','Oct',
                                          'Mar','Jul','Nov','Apr','Aug','Dec'),
                              "Year" = rep(2022,12),
                              "Max_Daily_Use_mgd" = as.numeric(Max_Day_Use)) %>%
  mutate(Month = recode(Month,
  Jan = 1,
  Feb = 2,
  Mar = 3,
  Apr = 4,
  May = 5,
  Jun = 6,
  Jul = 7,
  Aug = 8,
  Sep = 9,
  Oct = 10,
  Nov = 11,
  Dec = 12
)) %>% 
  mutate(Date = my(paste(Month,Year)))
max_daily_22_df %>% 
    arrange(ydm(Date))

#5 
max_daily_22_plot <- ggplot(max_daily_22_df,
                            aes(x=Date,
                                y=Max_Daily_Use_mgd))+ 
  geom_line()+
  labs(y="Max Daily Use (MGD)",
       title="Durham 2022 Maximum Daily Water Withdrawls")

max_daily_22_plot

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
#Create scraping function
Scrape_NC_DEQ <-function(year,PWSID){
  #Retrieve the website contents 
  website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=',
                              PWSID,'&year=',year))
  #Create element variables and scrape the webpage
  the_water_system_name <-website %>% 
    html_nodes('div+ table tr:nth-child(1) td:nth-child(2)')%>% 
    html_text()
  the_PWSID_code <-website %>% 
    html_nodes('td tr:nth-child(1) td:nth-child(5)')%>% 
    html_text()
  the_ownership <-website %>% 
    html_nodes('div+ table tr:nth-child(2) td:nth-child(4)')%>%
    html_text()
  the_max_day_use <-website %>% 
    html_nodes('th~ td+ td') %>% 
    html_text()
  #Convert to a dataframe and tidy the date columns
  max_daily_df_year <- data.frame("Water_System_Name" = rep(the_water_system_name,12),
                             "PWSID" = rep(the_PWSID_code,12),
                             "Ownership" = rep(the_ownership,12),
                             "Month" =c('Jan','May','Sep','Feb','Jun','Oct',
                                        'Mar','Jul','Nov','Apr','Aug','Dec'),
                             "Year" = rep(year,12),
                             "Max_Daily_Use_mgd" = as.numeric(the_max_day_use))%>%
    mutate(Month = recode(Month,
                          Jan = 1,
                          Feb = 2,
                          Mar = 3,
                          Apr = 4,
                          May = 5,
                          Jun = 6,
                          Jul = 7,
                          Aug = 8,
                          Sep = 9,
                          Oct = 10,
                          Nov = 11,
                          Dec = 12
                          )) %>% 
    mutate(Date = my(paste(Month,year))) %>% 
    arrange(ydm(Date))
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 
#Fetch Durham 2015
Durham_2015 <- Scrape_NC_DEQ('2015','03-32-010')

Durham_2015
```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 
#Fetch Asheville 2015
Asheville_2015 <- Scrape_NC_DEQ('2015','01-11-010')

Asheville_2015

#Create joint dataframe
Withdrawls_2015_df <-rbind(Durham_2015,Asheville_2015)

#Create joint plot
Withdrawls_2015_plot <-ggplot(Withdrawls_2015_df,
                              aes(x=Date,
                                  y=Max_Daily_Use_mgd,
                                  color=Water_System_Name))+ 
  geom_line()+
  labs(y="Max Daily Use (MGD)",
       title="2015 Maximum Daily Water Withdrawls")

Withdrawls_2015_plot

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2021.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 
#Subset the Asheville years
the_years <-  rep(2010:2021)

#Create a list of the PWSID we want, the same length as the vector above
the_PWSID <- rep.int('01-11-010',length(the_years))

#"Map" the "scrape.it" function to retrieve data for all these
Asheville_Max_Daily_DF <- map2(the_years,the_PWSID,Scrape_NC_DEQ) %>% bind_rows()

Asheville_Max_Daily_DF

#Plot
Asheville_Max_Daily_Plot <-ggplot(Asheville_Max_Daily_DF,
                                  aes(x=Date,
                                      y=Max_Daily_Use_mgd))+ 
  geom_line()+
  geom_smooth(method="loess",
              se=FALSE)+
  labs(y="Max Daily Use (MGD)",
       title="Asheville Maximum Daily Water Withdrawls 2010-2021")

Asheville_Max_Daily_Plot
```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer:
Asheville had a negative trend in water withdrawls between 2010-2015 but beginning in 2015 has a positive trend. This means they were using progressively less water between 2010-2015 but have been using an increasing amount between 2015-2021.
>
